Boosted Regression Trees for Data-Driven Auto-Tuning

### 2011-12-09 paper/draft1.tex
paper

\begin{abstract}
Auto-tuning is a widely used and effective technique for optimizing a
parametrized GPU code template for a particular computation on particular
hardware.  Its drawback is that thorough or exhaustive auto-tuning requires
compiling many kernels and calling each one many times; this process is slow.
Furthermore, library abstraction boundaries provide operations such as image
filtering and matrix multiplication, which actually correspond to a large set
of potential problem configurations with a wide variety of memory access
patterns and computational bottlenecks.  How can we draw on data from previous
empirical auto-tuning of related problems on related hardware to make a just-in-time
implementation decision for a novel problem?  This paper presents a machine
learning approach to auto-tuning, in which features of (a) the current
hardware platform, (b) the kernel configuration and (c) the problem instance
are passed to a regression model (boosted regression trees) which predicts how
much faster this kernel will be than a reference baseline.  Combinatorial
optimization strategies that would normally implement auto-tuning by
evaluating kernel configurations on the real hardware are orders of magnitude
faster when evaluating the regression model instead.  We validate our approach
using the filterbank correlation kernel described in \citet{pinto+cox:2011gcg}, where we find
that 0.1 seconds (XXX) of hill climbing on the regression model can achieve an
average of 95\% of the speed brought by 120 seconds of true auto-tuning XXX.  Our
approach is not specific to filterbank correlation, or even to GPU
auto-tuning: the approach of using a non-linear regression model on top of
simple features applies to a variety of problem types, kernel types, and
platforms.
\end{abstract}

@@

@@

\begin{figure}[!ht]
\centerline{
\begin{algorithm}{$\Algo{fbcorr}\big(gX, cF, gZ \big)$}
\Aitem shared $sX \setto$ all channels of region ($\beta$) of $gX$
\Aitem $x, y \setto$ position of this thread in output image
\Aitem \_\_syncthreads()
\Aitem $v[0:N] \setto 0$, for $N=4\times n\_output\_4s$
\Aitem for $d \setto 0$~\To~$D$,
\Aitem ~ for $h \setto 0$~\To~$H / n\_filter\_r$,
\Aitem ~ ~ for $w \setto 0$~\To~$W$,
\Aitem ~ ~ ~ $u \setto sX[x+h, y+w, d]$
\Aitem ~ ~ ~ for $n \setto 0$~\To~$n\_output\_4s - 1$,
\Aitem ~ ~ ~ ~  $v[n] \setto v[n] + cF[n, h, w, d]$
\Aitem for $n \setto 0$~\To~$n\_output\_4s - 1$,
\Aitem ~  gZ[x][y][4n:4n+n] = v[4n:4n+n], (float4)
\end{algorithm}
}
\caption{Kernel pseudo-code for filterbank correlation.
Input $gX$ is a pointer to $x$ in global memory, input $cF$ is a pointer to
$f$ in either constant or texture memory, and output $gZ$ is a pointer to $z$
in global memory.
Each block of threads modifies $4 \times n\_output\_4s$ channels of a
rectangle (called $\beta$ in code listing) within $z$. A grid of blocks covers all rows and columns of $z$.
Multiple calls can be used to apply all filters of a large filterbank $f$ to
$x$.
}
\label{fig:kernel}
\end{figure}

@@

The block height (block\_h) and block width (block\_w) parameters control the
number of threads that run on each multiprocessor.
The number of filter rows (n\_filter\_r) is how many rows of each filter is
loaded

>>>>

\begin{abstract}
Auto-tuning is a widely used and effective technique for optimizing a
parametrized GPU code template for a particular computation on particular
hardware.  Its drawback is that thorough or exhaustive auto-tuning requires
compiling many kernels and calling each one many times; this process is slow.
Furthermore, library abstraction boundaries provide operations such as image
filtering and matrix multiplication, which actually correspond to a large set
of potential problem configurations with a wide variety of memory access
patterns and computational bottlenecks.  How can we draw on data from previous
empirical auto-tuning of related problems on related hardware to make a just-in-time
implementation decision for a novel problem?  This paper presents a machine
learning approach to auto-tuning, in which features of (a) the current
hardware platform, (b) the kernel configuration and (c) the problem instance
are passed to a regression model (boosted regression trees) which predicts how
much faster this kernel will be than a reference baseline.  Combinatorial
optimization strategies that would normally implement auto-tuning by
evaluating kernel configurations on the real hardware are orders of magnitude
faster when evaluating the surrogate regression model instead.  We validate our approach
using the filterbank correlation kernel described in \citet{pinto+cox:2011gcg}, where we find
that 0.1 seconds (XXX) of hill climbing on the regression model ({\em
predictive}
auto-tuning) can achieve an
average of 95\% of the speed brought by 120 seconds of empirical auto-tuning XXX.  Our
approach is not specific to filterbank correlation, or even to GPU
kernel auto-tuning: the approach of using a non-linear regression model on top of
simple features applies to a variety of problem types, kernel types, and
platforms.
\end{abstract}

@@

In terms of floating point operations, a filterbank correlation requires the
inner sums to be computed for each output pixel, yielding the quantity in
Eq.~\ref{eq:flops}:
\begin{equation}
\mathrm{FLOPS} = 2  K  H  W  D  (R - H + 1)( C- W + 1)
\label{eq:flops}
\end{equation}
The multiplicative factor of 2 arises because we must multiply an element of
$x$ with an element of $f$ and then add the result to an element of $z$.

In terms of bandwidth, if we were to assume (optimistically) that the entire
filterbank fit into the GPU's so-called constant memory, then the bandwidth
required for a filterbank correlation would be the size of the filterbank,
plus the size of the input image, plus the size of the output image. This sum
is expressed in terms of configuration parameters in Eq.~\ref{eq:bytes}:
\begin{align}
\mathrm{Bytes} = 4&RCD \nonumber \\
& + 4KHWD \nonumber \\
& + 4(R-H+1)(C-W+1)K
\label{eq:bytes}
\end{align}
We will see the bandwidth requirements of a more realistic analysis in
Section~XXX.

@@

\begin{figure}[!ht]
\centerline{
\begin{algorithm}{$\Algo{fbcorr}\big(gX, cF, gZ \big)$}
\Aitem shared $sX \setto$ all channels of region ($\beta$) of $gX$
\Aitem $x, y \setto$ position of this thread in output image
\Aitem \_\_syncthreads()
\Aitem $v[0:N] \setto 0$, for $N=4\times n\_output\_4s$
\Aitem for $d \setto 0$~\To~$D$,
\Aitem ~ for $h \setto 0$~\To~$H / n\_filter\_r$,
\Aitem ~ ~ for $w \setto 0$~\To~$W$,
\Aitem ~ ~ ~ $u \setto sX[x+h, y+w, d]$
\Aitem ~ ~ ~ for $n \setto 0$~\To~$n\_output\_4s - 1$,
\Aitem ~ ~ ~ ~  $v[n] \setto v[n] + cF[n, h, w, d]$
\Aitem for $n \setto 0$~\To~$n\_output\_4s - 1$,
\Aitem ~  gZ[x][y][4n:4n+n] += v[4n:4n+n], (float4)
\end{algorithm}
}
\caption{Kernel pseudo-code for filterbank correlation.
Input $gX$ is a pointer to $x$ in global memory, input $cF$ is a pointer to
$f$ in either constant or texture memory, and output $gZ$ is a pointer to $z$
in global memory.
Each block of threads modifies $4 \times n\_output\_4s$ channels of a
rectangle (called $\beta$ in code listing) within $z$. A grid of blocks covers all rows and columns of $z$.
Multiple calls can be used to apply all filters of a large filterbank $f$ to
$x$.
}
\label{fig:kernel}
\end{figure}

@@

The block height (``block\_h'') and block width (``block\_w'') parameters control the
number of threads that run within each block.
Each kernel call loads some number of filter rows (``n\_filter\_r'') into
constant memory and processes the correlation of the image with just those
rows, incrementing the output buffer.
Each thread can compute several output elements
at once, in multiples (``n\_output\_4s'') of 4;
this increases the efficiency of each thread, but can lead to lower occupancy.
Registers are a precious commodity on the GPU, and this kernel accumulates
elements of $z$ in registers.
The ``spill'' parameter controls whether the current thread's output position in
$z$ is stored in a register (faster access) or in shared memory (frees up
a register).
The ``imul\_fast'' parameter controls whether integer multiplication is done
in 24-bit (True) or 32-bit (False) precision.
The ``pad\_shared'' parameter controls whether the $sX$ shared memory buffer
is padded, which wastes space in shared memory but reduces bank conflicts.
The ``use\_tex1d'' parameter controls whether the image is loaded into shared
memory with global pointer dereferences or texture fetches.
The ``maxrreg'' and ``fast\_math'' parameters are passed to the nvcc compiler to limit
the number of registers available to each thread, and to enable XXX,
respectively.
%XXX: what effect does fast-math have here?

Often however, the entire filterbank often does not fit into the GPU's
constant memory and so $P$ passes are necessary to compute $z$. In such cases,
the number of bytes moved is scaled almost linearly in $P$:
\begin{align}
\mathrm{Bytes} = 4&RCDP \nonumber \\
& + 4KHWD \nonumber \\
& + 8(R-H+1)(C-W+1)KP
% XXX: incrementing requires reading and writing $z$ right? so actually the 4
% above should be an 8? (see related ticket on github for potential
% optimization about this)
\label{eq:bytesP}
\end{align}

Filterbank correlation represents a difficult code-optimization problem
because of the number of interacting constraints: a large value of $P$ reduces
arithmetic density, but a small value of $P$ requires that each thread do more
work, and makes global memory access latency harder to hide. (XXX: is there a
better reason?)
